---
title: "Regression Models Project"
author: "AIS1209"
date: "December 22, 2015"
output: pdf_document
fig_caption: true
number_section: true
fontsize: 10
---
# Executive Summary
*Motor Trend*, a magazine about the automobile industry, is interested in exploring the relationship between a set of variables and miles per gallon (MPG, outcome) in a data set of a collection of cars. This project will (1) determine whether automatic or manual transmission is better for MPG, and (2) quantify the MPG difference between automatic and manual transmissions. In order to accomplish these goals, several linear models were fitted, and it was found that while mpg depends on several variables, only one model could be used to answer determine which of the two transmission types prodces better mileage and to provide a numeric value for the difference in the mean mpg values.

# Exploratory Data Analysis
The data consists of the `mtcars` data set available in R. It consists of the following variables
```{r, echo= FALSE}
library(datasets); library(lattice); library(ggplot2); library(knitr); library(car)
data(mtcars)
names(mtcars)
```
Variable mpg is selected as the outcome and variable am is an obvious candidate for the predictor variable list. An exploratory analysis of the data was performed in the form of a correlation matrix and some exploratory plots of various variables in the data set.  

The exploratory analysis shown in the Appendix (correlation matrix and plots) suggest that (1) there appears to be a relationship between mpg and transmission type, (2) there is significant correlation, as indicated by correlation coefficients, among some of the variables, and (3) in order to quantify the mpg difference between automatic and manual transmission, several other variables may have to be taken into account. For example, the exploratory plots suggest that mpg is higher for manual cars than for automatic cars in the case of 4 cylinders, but not in the case of 6 or 8 cylinders.

# Model Fitting
First, a simple model of the form $mpg = \beta_0 + \beta_1 (am)$ was fitted, where am is codes as a factor variable.
```{r, echo = FALSE}
mtcars1 <- mtcars
mtcars <- transform(mtcars, am = factor(am), cyl = factor(cyl), 
                    gear = factor(gear), carb = factor(carb), vs = factor(vs))
fit1 <- lm(mpg ~ am, data = mtcars)
summary(fit1)$coefficients
```
Both regression coeficients are significant, according to their p-values; however the $R^2$ coefficient is quite low at 0.3598 (see discussion in the next section). According to this model, the intercept represents the mean mpg for automatic transmission cars = 17.147 mpg; the slope + intercept = 24.392 mpg represents the mean of the mpg values for manual transmission cars. Therefore the model suggests that manual cars are better than automatic ones in terms of mpg. Given that the exploratory analysis suggests an interrelationship among variables, a nested model fitting was attempted, by iteratively adding variables wt, cyl, hp, gear, carb and vs as factors. An ANOVA analysis suggests that wt, cyl6, hp are significant for the model, but adding them makes the variable am not significant. Note that cyl8, gear, and carb are not significant (large p-values). Another analysis (not shown) determined that disp does not add any value to the model, and neither do interaction terms.
```{r, echo = F}
fit2 <- lm(mpg ~ am + wt, data = mtcars)
fit3 <- lm(mpg ~ am + wt + cyl, data = mtcars)
fit4 <- lm(mpg ~ am + wt + cyl + hp, data = mtcars)
fit5 <- lm(mpg ~ am + wt + cyl + hp+ gear, data = mtcars)
fit6 <- lm(mpg ~ am + wt + cyl + hp + gear + carb , data = mtcars)
fit7 <- lm(mpg ~ am + wt + cyl + hp + gear + carb + vs , data = mtcars)
anova(fit1, fit2, fit3, fit4, fit5, fit6, fit7)
# mdl <- update(mdl, ~.+ disp); summary(mdl); mdl <- update(mdl, ~. -disp)
#fit4_1 <- lm(mpg ~ am * wt * cyl * hp, data = mtcars); anova(fit4, fit4_1)
```

Model $mpg ~ am + wt + cyl + hp$:
```{r, echo = FALSE}
mdl <- fit4
summary(fit4)$coefficients
```

# Diagnostic Plots and Discussion of Uncertainty/Inference
As discussed in the previous section, the residuals plots indicate a less than optimal fit for the simple model with only transmission type as predictor. The residual distribution appears closed to a normal one in the case of the second model. It appears that variables other than transmission type also influence the mpg; the goodness-of-fit $R^2$ coefficients, also indicate that the second model is a better description of what factor produce a better mpg value.
```{r, echo = FALSE}
df <- data.frame(Rsq_fit_am_only = summary(fit1)$r.squared, Rsq_fit_multiple = summary(fit4)$r.squared)
print(df)
```

```{r, echo = FALSE, fig.height= 4}
par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
plot(fit1, which = 2, main = "mpg ~ am", sub.caption = "Normal probability plots")
plot(fit4, which = 2, main = "mp ~ am + wt + cyl + hp", sub.caption = "")
```

However, in the second model, the large p-value corresponding to the coefficient for the factor variable am indicates that the variable may not be necessary for the model (the null hypothesis that the coefficent of am is zero is accepted). This inference is also supported by the 95% confidence interval for this coefficient, which includes zero. Therefore, the second model cannot be used to quantify the difference in mpg values between automatic and manual transmission. The first model (fit shown in the Appendix) predicts that the mean mpg value is 7.24 mpg higher in manual cars than in automatic cars.
```{r, echo = FALSE}
# The NULL hypothesis is that your am coefficient is not different to zero. The
# p value is an indication of the likelihood of getting the am coefficient value
# that you did if the NULL hypothesis is true. That is, how likely is it that I
# would have got this am coefficient (or higher) if the value is actually zero.
confint(fit4)[2, 1:2]
```


# Appendix
## Correlation matrix
```{r, echo = FALSE}
kable(round(cor(mtcars1),2), caption = "Correlation matrix for the mtcars dataset")
```

## Exploratory Plots
```{r, echo=FALSE, fig.cap= "mpg vs. transmission type", fig.height= 3, fig.width=3}
plot(mtcars$am, mtcars$mpg, xlab = "Transmission", ylab = "mpg", main = "mpg vs. transmission")
q <- ggplot(mtcars, aes(am, mpg))
q +facet_grid(. ~ cyl) + ylab("Cylinders") + labs(title = 'Cylinders') + xlab("Transmission")+geom_point()
q <- ggplot(mtcars, aes(am, mpg))
q +facet_grid(. ~ vs) + labs(title = 'Variable Speed') + xlab("Transmission") + geom_point()
q <- ggplot(mtcars, aes(am, mpg))
q +facet_grid(. ~ gear) + labs(title = 'Gear') + xlab("Transmission") + geom_point()
```

## Model fit lm(mpg ~ am)
```{r, echo = FALSE, fig.height=3}
#xyplot(mpg ~ am, data = mtcars, layout = c(1, 1), main = "mpg vs. transmission type")
plot(mtcars1$am, mtcars1$mpg, xlab = "transmission", ylab = "mpg")
abline(lm(mpg ~ am, data=mtcars), lwd = 2)
```

## Code
```{r, eval=F, tidy = F}
library(datasets); library(lattice); library(ggplot2); library(knitr); library(car)
data(mtcars); names(mtcars); mtcars1 <- mtcars
mtcars <- transform(mtcars, am = factor(am), cyl = factor(cyl), gear = factor(gear), 
                    carb = factor(carb), vs = factor(vs))
fit1 <- lm(mpg ~ am, data = mtcars)
summary(fit1)$coefficients
fit2 <- lm(mpg ~ am + wt, data = mtcars)
fit3 <- lm(mpg ~ am + wt + cyl, data = mtcars)
fit4 <- lm(mpg ~ am + wt + cyl + hp, data = mtcars) # use this model
fit5 <- lm(mpg ~ am + wt + cyl + hp+ gear, data = mtcars)
fit6 <- lm(mpg ~ am + wt + cyl + hp + gear + carb , data = mtcars)
anova(fit1, fit2, fit3, fit4, fit5, fit6)
# mdl <- update(mdl, ~.+ disp); summary(mdl); mdl <- update(mdl, ~. -disp)
#fit4_1 <- lm(mpg ~ am * wt * cyl * hp, data = mtcars); anova(fit4, fit4_1)
summary(fit4)$coefficients # Keep fit4
summary(fit4)$r.squared
par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
plot(fit1, which = 2, main = "mpg ~ am", sub.caption = "Normal probability plots")
plot(fit4, which = 2, main = "mp ~ am + wt + cyl + hp", sub.caption = "")
print(data.frame(Rsq_fit_am_only = summary(fit1)$r.squared, 
                 Rsq_fit_multiple = summary(fit4)$r.squared))
confint(fit4)[2, 1:2]
plot(mtcars$am, mtcars$mpg, xlab = "Transmission", ylab = "mpg", main = "mpg vs. transmission")
q <- ggplot(mtcars, aes(am, mpg)) + facet_grid(. ~ cyl) + ylab("Cylinders") + labs(title = 'Cylinders') + xlab("Transmission")+geom_point()
q <- ggplot(mtcars, aes(am, mpg)) + facet_grid(. ~ vs) + labs(title = 'Variable Speed') + 
    xlab("Transmission") + geom_point()
q <- ggplot(mtcars, aes(am, mpg)) + facet_grid(. ~ gear) + labs(title = 'Gear') + 
    xlab("Transmission") + geom_point()
plot(mtcars1$am, mtcars1$mpg, xlab = "transmission", ylab = "mpg"); abline(lm(mpg ~ am, data=mtcars))
```
